%=====================================================================
% FILE: doc/latex/flux_and_spectrum_definitions.tex
% INTENT:
%   LaTeX source for a documentation PDF linked from the in-app Docs
%   menu. The PDF is intended to be read alongside the wizard and
%   provide paper-style detail: definitions, equations, references,
%   and worked examples.
%
% METHODS / DESIGN:
%   - Keep structure consistent across docs: Abstract → Background →
%     Methods → Implementation Notes → Example(s) → References.
%   - Use clear variable definitions and units.
%   - Prefer short, self-contained equations over long derivations.
%
% IMPLEMENTATION NOTES:
%   - Generated PDFs live in doc/*.pdf; sources are in doc/latex/*.tex.
%   - Keep equations compatible with pdflatex.
%
% LAST UPDATED: 2026-03-14
%=====================================================================
% ==============================================================================
% FILE:    flux_and_spectrum_definitions.tex
% PROJECT: AMPS CCMC SEP/GCR Transport Interface
% PURPOSE: Definitions of flux / intensity / spectra and unit conventions,
%          including the gridless density/spectrum solver and its dipole
%          analytic comparison benchmarks.
%
% BUILD:
%   pdflatex -interaction=nonstopmode -halt-on-error flux_and_spectrum_definitions.tex
% OUTPUT:
%   ../flux_and_spectrum_definitions.pdf
% ==============================================================================
\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,bm}
\usepackage{hyperref}
\usepackage{siunitx}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{array}
\usepackage{longtable}
\setlist{nosep}
\hypersetup{colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue}

\title{Flux, Spectrum, Density, and Anisotropic Boundary Conditions Used by AMPS}
\author{AMPS Interface Documentation}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This note documents the quantities used by the AMPS gridless energetic-particle
post-processing tools for cutoff rigidity, local spectrum, omnidirectional flux,
and number density. It also describes how isotropic and anisotropic boundary
spectra are implemented in the current code, and how the internal dipole-field
reference solutions used for nightly tests are constructed for comparison with
numerical model output in both \texttt{POINTS} and \texttt{SHELLS} modes.
\end{abstract}

\section{Differential intensity and related quantities}
Throughout the AMPS interface, the particle spectrum is specified as a
\emph{differential directional intensity}
\begin{equation}
  J(E,\Omega) \equiv \frac{dN}{dA\,dt\,d\Omega\,dE},
\end{equation}
with typical units
\begin{equation}
  J(E,\Omega)\;[\text{particle}~\text{m}^{-2}\,\text{s}^{-1}\,\text{sr}^{-1}\,\text{MeV}^{-1}].
\end{equation}
The boundary spectrum supplied to the gridless density/spectrum tool is a model
for this quantity at the outer boundary of the tracing domain.

\paragraph{Energy variable.}
For ions, AMPS often uses kinetic energy per nucleon in \si{MeV/nucleon} in the
user interface. Internally, the density/spectrum integration is carried out over
an energy grid $E_i$ in \si{MeV}; the same spectral model is queried at each
energy point and folded with geomagnetic transmissivity.

\paragraph{Omnidirectional differential flux.}
The omnidirectional differential flux is the solid-angle integral of the
intensity,
\begin{equation}
  j(E) = \int_{4\pi} J(E,\Omega)\,d\Omega.
\end{equation}
If the distribution is isotropic, then
\begin{equation}
  j(E)=4\pi J(E).
\end{equation}
In anisotropic cases the factor $4\pi$ cannot be inserted by hand; the angular
integral must be evaluated with the actual angular weighting.

\paragraph{Integral flux and pfu.}
The omnidirectional integral flux above threshold $E_0$ is
\begin{equation}
  F(>E_0)=\int_{E_0}^{\infty} j(E)\,dE,
\end{equation}
or, equivalently, in isotropic form,
\begin{equation}
  F(>E_0)=4\pi\int_{E_0}^{\infty} J(E)\,dE.
\end{equation}
NOAA/SWPC often reports proton event fluxes in pfu, where
\begin{equation}
  1~\text{pfu}=1~\text{proton}~\text{cm}^{-2}\,\text{s}^{-1}\,\text{sr}^{-1}
\end{equation}
for an \emph{integral} channel. When fitting a model to integral channels, AMPS
computes the energy integral of the chosen spectral model.

\section{From intensity to local spectrum, flux, and density}
The gridless solver does not solve a full transport equation in phase space.
Instead, for each evaluation location $\bm{x}_0$ and energy $E$, it computes the
fraction of directions at the outer boundary that can access that point by
backtracing particles in the geomagnetic field. This fraction is the
\emph{transmissivity} $T(E,\bm{x}_0)$.

\subsection{Boundary spectrum and local spectrum}
Let $J_b(E)$ denote the boundary differential intensity model at the outer
boundary. In the isotropic case the local differential intensity is
\begin{equation}
  J_{\mathrm{loc}}(E;\bm{x}_0)=T(E,\bm{x}_0)\,J_b(E).
  \label{eq:Jloc_iso}
\end{equation}
Thus the magnetosphere acts as an energy- and location-dependent filter on the
specified boundary spectrum.

\subsection{Number density from intensity}
For an isotropic distribution, the differential number density $n_E(E)$ and
intensity are related by
\begin{equation}
  n_E(E)=\frac{4\pi}{v(E)}J(E),
\end{equation}
where $v(E)$ is the particle speed corresponding to kinetic energy $E$. The
number density is therefore
\begin{equation}
  n(\bm{x}_0)=4\pi\int \frac{J_{\mathrm{loc}}(E;\bm{x}_0)}{v(E)}\,dE.
  \label{eq:density_def}
\end{equation}
This factor $1/v(E)$ is physically important: low-energy particles contribute
more strongly to density than to integral flux because they linger longer per
unit path length.

\subsection{Omnidirectional integral flux}
The total omnidirectional integral flux over the modeled energy band is
\begin{equation}
  F(\bm{x}_0)=4\pi\int J_{\mathrm{loc}}(E;\bm{x}_0)\,dE.
  \label{eq:flux_def}
\end{equation}
For a user-defined channel $[E_1,E_2]$, the channel flux is
\begin{equation}
  F_{[E_1,E_2]}(\bm{x}_0)=4\pi\int_{E_1}^{E_2} J_{\mathrm{loc}}(E;\bm{x}_0)\,dE.
  \label{eq:flux_channel_def}
\end{equation}
The code evaluates Eqs.~\eqref{eq:density_def}--\eqref{eq:flux_channel_def}
with trapezoidal quadrature on the same energy grid used for transmissivity.

\subsection{Relativistic conversion between energy and speed}
Let $m_0$ be the particle rest mass and $T$ the kinetic energy in joules. The
Lorentz factor is
\begin{equation}
  \gamma = 1 + \frac{T}{m_0 c^2},
\end{equation}
and the speed is
\begin{equation}
  v(E)=c\sqrt{1-\frac{1}{\gamma^2}}.
\end{equation}
This relativistic speed is used consistently when converting local intensity to
number density.

\section{Isotropic and anisotropic boundary spectra}
\subsection{Isotropic boundary mode}
If the boundary distribution is isotropic, the boundary intensity depends only on
energy,
\begin{equation}
  J_b(E,\Omega)=J_b(E).
\end{equation}
In that case the only angular dependence entering the local solution is the
geomagnetic accessibility. The code samples a finite direction set
$\{\Omega_k\}_{k=1}^{N_\Omega}$ and computes
\begin{equation}
  T(E,\bm{x}_0) \approx \frac{1}{N_\Omega}\sum_{k=1}^{N_\Omega} I_k(E,\bm{x}_0),
  \label{eq:T_iso_discrete}
\end{equation}
where $I_k=1$ if the reversed trajectory launched from $\bm{x}_0$ in direction
$\Omega_k$ escapes the outer boundary and $I_k=0$ otherwise.

Substituting Eq.~\eqref{eq:T_iso_discrete} into Eq.~\eqref{eq:Jloc_iso} yields
the discrete isotropic approximation implemented in the code.

\subsection{Anisotropic boundary mode}
If the boundary distribution is anisotropic, the code models it as an isotropic
spectral shape multiplied by a directional and/or spatial weighting factor. In
that case the boundary intensity is written schematically as
\begin{equation}
  J_b(E,\Omega,\bm{x}_{\mathrm{exit}})
  = J_b(E)\,f_{\mathrm{aniso}}(\Omega,\bm{x}_{\mathrm{exit}}),
\end{equation}
where $\bm{x}_{\mathrm{exit}}$ is the position where the backtraced trajectory
reaches the outer boundary. In the present implementation the anisotropy factor
is separated into
\begin{equation}
  f_{\mathrm{aniso}} = f_{\mathrm{PAD}}(\mu)\,f_{\mathrm{spatial}}(\bm{x}_{\mathrm{exit}}),
  \qquad \mu = \cos\alpha,
  \label{eq:aniso_factorization}
\end{equation}
where $\alpha$ is the pitch angle with respect to the local magnetic field at the
exit point.

The code therefore computes a weighted transmissivity
\begin{equation}
  T_{\mathrm{aniso}}(E,\bm{x}_0)
  \approx
  \frac{\sum_{k=1}^{N_\Omega} I_k(E,\bm{x}_0)\,w_k}
       {\sum_{k=1}^{N_\Omega} w_k},
  \qquad w_k = f_{\mathrm{PAD}}(\mu_k)\,f_{\mathrm{spatial}}(\bm{x}_{\mathrm{exit},k}).
  \label{eq:T_aniso_discrete}
\end{equation}
If all weights are unity, Eq.~\eqref{eq:T_aniso_discrete} reduces to the
isotropic average in Eq.~\eqref{eq:T_iso_discrete}.

\subsection{Physics of the anisotropy model}
The anisotropic option is meant to represent the physically common situation in
which the outer-boundary population is not uniformly distributed in direction or
space. Examples include:
\begin{itemize}
  \item field-aligned or anti-field-aligned streaming following solar injection,
  \item bidirectional populations produced by magnetic mirroring or trapping,
  \item day-night or regional asymmetries in the outer magnetosphere, and
  \item user-prescribed pitch-angle distributions for sensitivity studies.
\end{itemize}
The purpose of the implementation is not to solve a self-consistent kinetic
transport problem at the boundary, but to apply a physically transparent weighting
model to the trajectories that successfully connect the outer boundary to the
observation point.

\subsection{Pitch-angle distribution models}
The code supports several pitch-angle weighting laws through
\texttt{EvalAnisotropyFactor}:
\begin{itemize}
  \item \texttt{ISOTROPIC}: $f_{\mathrm{PAD}}=1$;
  \item \texttt{SINALPHA\_N}: $f_{\mathrm{PAD}}=(\sin\alpha)^N=(1-\mu^2)^{N/2}$;
  \item \texttt{COSALPHA\_N}: $f_{\mathrm{PAD}}=|\cos\alpha|^N=|\mu|^N$;
  \item \texttt{BIDIRECTIONAL}: a symmetric enhancement about $\mu=\pm 1$.
\end{itemize}
These provide a simple way to emulate broad classes of distributions ranging from
quasi-perpendicular populations to strongly field-aligned beams.

\subsection{Spatial weighting models}
The present implementation also permits a multiplicative spatial weighting,
$ f_{\mathrm{spatial}}(\bm{x}_{\mathrm{exit}})$, at the exit point of the
backtraced trajectory. For example, a day-night prescription can assign different
weights on the dayside and nightside outer boundary. The physical interpretation
is that the intensity incident on the outer boundary is not spatially uniform,
so trajectories connecting to favored boundary regions contribute more strongly to
the local population.

\subsection{Local spectrum, density, and flux in anisotropic mode}
Once the weighted transmissivity of Eq.~\eqref{eq:T_aniso_discrete} is obtained,
the remaining formulas are unchanged:
\begin{align}
  J_{\mathrm{loc}}(E;\bm{x}_0) &= T_{\mathrm{aniso}}(E,\bm{x}_0)\,J_b(E), \\
  n(\bm{x}_0) &= 4\pi\int \frac{J_{\mathrm{loc}}(E;\bm{x}_0)}{v(E)}\,dE, \\
  F(\bm{x}_0) &= 4\pi\int J_{\mathrm{loc}}(E;\bm{x}_0)\,dE.
\end{align}
Thus anisotropy changes the \emph{effective transmission factor}, not the
fundamental definitions of spectrum, density, or flux.

\section{Numerical algorithm implemented in the gridless solver}
For either \texttt{POINTS} or \texttt{SHELLS} mode, the density/spectrum solver
follows the same algorithm.

\begin{enumerate}
  \item Build the requested energy grid $\{E_i\}$, either linearly or
        logarithmically spaced.
  \item Build the discrete direction set $\{\Omega_k\}$ used to sample arrival
        directions at the evaluation point.
  \item For each location $\bm{x}_0$ and energy $E_i$, convert energy to rigidity
        $R_i$ and backtrace reversed trajectories in the prescribed magnetic field.
  \item For each sampled direction, determine whether the trajectory reaches the
        outer boundary (allowed) or intersects the inner loss sphere / fails the
        termination criterion (forbidden).
  \item Form the transmissivity using either the unweighted isotropic average,
        Eq.~\eqref{eq:T_iso_discrete}, or the weighted anisotropic average,
        Eq.~\eqref{eq:T_aniso_discrete}.
  \item Evaluate the boundary spectrum $J_b(E_i)$ and form the local spectrum
        $J_{\mathrm{loc}}(E_i)=T(E_i)J_b(E_i)$.
  \item Integrate the local spectrum over energy to obtain total number density,
        total omnidirectional integral flux, and any requested energy-channel fluxes.
  \item Write output files containing the total scalar fields and, where requested,
        the full transmissivity and local spectrum as functions of energy.
\end{enumerate}

\paragraph{POINTS mode.}
In \texttt{POINTS} mode the tool evaluates a user-supplied list of locations and
writes pointwise files for density, spectrum, and flux.

\paragraph{SHELLS mode.}
In \texttt{SHELLS} mode the tool evaluates a structured longitude-latitude grid on
one or more altitude shells and writes the corresponding shell maps. The local
physics and energy integration are identical to \texttt{POINTS} mode; only the
sampling of $\bm{x}_0$ differs.

\section{Analytic and semi-analytic dipole reference solutions used for nightly tests}
To validate the implementation, the code contains internal reference calculations
for the special case of a dipole geomagnetic field. These references are written
to comparison files such as
\begin{center}
\texttt{density\_gridless\_points\_dipole\_compare.dat},\quad
\texttt{spectrum\_gridless\_points\_dipole\_compare.dat},\quad
\texttt{flux\_gridless\_points\_dipole\_compare.dat}
\end{center}
and their \texttt{SHELLS} analogs when nightly-test mode is enabled.

\subsection{Purpose of the dipole benchmark}
The benchmark is designed to test the entire chain
\begin{equation}
  \text{cutoff} \rightarrow \text{transmissivity} \rightarrow \text{local spectrum}
  \rightarrow \text{density / flux integrals},
\end{equation}
while using a field model simple enough that the dominant geometric behavior is
understood analytically. It is not intended to replace the full numerical model,
but to provide a stable and interpretable regression test.

\subsection{Vertical St\'ormer cutoff used in the reference}
For a dipole field, the classical St\'ormer theory provides an analytic cutoff
rigidity for special geometries. The internal benchmark uses the \emph{vertical}
St\'ormer cutoff,
\begin{equation}
  R_{c,\mathrm{vert}}(\lambda,r)
  \propto \frac{\cos^4\lambda}{r^2},
\end{equation}
where $\lambda$ is magnetic latitude and $r$ is radial distance in Earth radii.
The code evaluates this quantity through an internal helper and uses it as a
hard transition in rigidity space:
\begin{equation}
  H(E;\bm{x}_0)=
  \begin{cases}
    0, & R(E) < R_{c,\mathrm{vert}}(\bm{x}_0),\\
    1, & R(E) \ge R_{c,\mathrm{vert}}(\bm{x}_0).
  \end{cases}
  \label{eq:stormer_step}
\end{equation}
Equation~\eqref{eq:stormer_step} is the spectrum-space analogue of the usual
analytic dipole cutoff benchmark.

\subsection{Isotropic geometric factor in the reference}
The numerical solver uses curved trajectories in the dipole field. The internal
analytic reference intentionally uses a simpler geometry: a straight-line
``open-sky'' estimate for the fraction of directions from which a particle can
reach the outer boundary without intersecting the inner spherical absorber. Let
$G_{\mathrm{sky}}(\bm{x}_0)$ denote that purely geometric fraction of accessible
solid angle. Then the isotropic reference transmissivity is
\begin{equation}
  T_{\mathrm{ref}}^{\mathrm{iso}}(E,\bm{x}_0)
  = G_{\mathrm{sky}}(\bm{x}_0)\,H(E;\bm{x}_0).
  \label{eq:Tref_iso}
\end{equation}
This separates the problem into a geometric factor and an energy threshold.

\subsection{Anisotropic geometric factor in the reference}
For anisotropic nightly tests, the reference cannot be reduced to a single
constant open-sky factor because different accessible directions carry different
weights. The code therefore computes a \emph{semi-analytic} reference by
integrating the prescribed anisotropy weight over the straight-line accessible
solid angle:
\begin{equation}
  G_{\mathrm{sky}}^{\mathrm{aniso}}(\bm{x}_0)
  = \frac{\int_{\Omega \in \Omega_{\mathrm{open}}}
     f_{\mathrm{aniso}}(\Omega,\bm{x}_{\mathrm{exit}})\,d\Omega}
    {\int_{4\pi} f_{\mathrm{aniso}}(\Omega,\bm{x}_{\mathrm{exit}})\,d\Omega},
  \label{eq:Gsky_aniso}
\end{equation}
where $\Omega_{\mathrm{open}}$ is the subset of directions that do not intersect
the inner absorber in the straight-line construction. The anisotropic reference
transmissivity is then
\begin{equation}
  T_{\mathrm{ref}}^{\mathrm{aniso}}(E,\bm{x}_0)
  = G_{\mathrm{sky}}^{\mathrm{aniso}}(\bm{x}_0)\,H(E;\bm{x}_0).
  \label{eq:Tref_aniso}
\end{equation}
This benchmark is still much simpler than the full backtracing solution, but it
is sensitive to errors in anisotropy weighting, angular normalization, and the
subsequent spectrum-folding and integration steps.

\subsection{Reference local spectrum, density, and flux}
Once $T_{\mathrm{ref}}(E,\bm{x}_0)$ is known from either
Eq.~\eqref{eq:Tref_iso} or Eq.~\eqref{eq:Tref_aniso}, the reference local
spectrum is
\begin{equation}
  J_{\mathrm{loc,ref}}(E;\bm{x}_0)=T_{\mathrm{ref}}(E,\bm{x}_0)\,J_b(E).
\end{equation}
The reference density and flux are then computed by the \emph{same} formulas as
the numerical solution,
\begin{align}
  n_{\mathrm{ref}}(\bm{x}_0) &= 4\pi\int \frac{J_{\mathrm{loc,ref}}(E;\bm{x}_0)}{v(E)}\,dE, \\
  F_{\mathrm{ref}}(\bm{x}_0) &= 4\pi\int J_{\mathrm{loc,ref}}(E;\bm{x}_0)\,dE, \\
  F_{\mathrm{ref},[E_1,E_2]}(\bm{x}_0) &= 4\pi\int_{E_1}^{E_2} J_{\mathrm{loc,ref}}(E;\bm{x}_0)\,dE.
\end{align}
Using the same quadrature and the same boundary-spectrum evaluator in both the
reference and numerical branches isolates errors in geomagnetic filtering and
anisotropy handling from errors in spectral integration.

\subsection{What is compared in the output files}
The comparison files report, depending on the product:
\begin{itemize}
  \item numerical versus reference density,
  \item numerical versus reference transmissivity spectra,
  \item numerical versus reference local spectra, and
  \item numerical versus reference total and channel-integrated fluxes.
\end{itemize}
These outputs are intentionally separated because different bugs show up most
clearly in different products:
\begin{itemize}
  \item a cutoff or rigidity-conversion bug appears first in transmissivity and
        local spectrum,
  \item a speed-conversion bug appears most strongly in number density because of
        the factor $1/v(E)$, and
  \item a channel-integration or interpolation bug appears most strongly in the
        integrated flux outputs.
\end{itemize}

\section{Rigidity spectra and variable changes}
Some literature expresses spectra in rigidity $R$ instead of energy:
\begin{equation}
  J_R(R)=\frac{dN}{dA\,dt\,d\Omega\,dR}.
\end{equation}
The conversion between energy-space and rigidity-space spectra is
\begin{equation}
  J_R(R)=J_E(E)\left|\frac{dE}{dR}\right|,
\end{equation}
where the mapping $E(R)$ depends on particle mass and charge because
$R=pc/(Ze)$. The cutoff-rigidity solver and the density/spectrum solver therefore
share the same species-dependent energy-rigidity conversion.

\section{Implementation notes for the current AMPS code}
The present gridless density/spectrum tool uses the same trajectory classifier as
the gridless cutoff-rigidity tool. This is important for consistency: the density
solver does not maintain a separate magnetic-field evaluator, particle pusher, or
termination logic. Instead, both cutoff and density calculations rely on the same
single-particle accessibility test, and differ only in how they post-process the
result:
\begin{itemize}
  \item cutoff rigidity seeks the threshold in rigidity for a given location and
        direction set;
  \item density/spectrum computes transmissivity over an energy grid and folds it
        with a boundary spectrum.
\end{itemize}
This shared-core design reduces the risk that the two tools drift apart because
of subtle implementation differences.

\section{Summary}
The AMPS gridless density/spectrum model computes local energetic-particle
properties by filtering a specified boundary spectrum through geomagnetic
transmissivity. In isotropic mode, transmissivity is the unweighted fraction of
allowed directions; in anisotropic mode it is a weighted fraction reflecting the
prescribed pitch-angle and spatial dependence at the outer boundary. The local
spectrum is then integrated over energy to obtain omnidirectional flux and number
density. For the dipole field, AMPS includes built-in analytic and semi-analytic
reference solutions based on the vertical St\'ormer cutoff and a straight-line
open-sky geometry, making it possible to compare numerical and reference results
for density, spectrum, and flux in both \texttt{POINTS} and \texttt{SHELLS} modes
without relying on external scripts.

\section{References (selected)}
\begin{itemize}
  \item Grieder, P. K. F. (2001), \emph{Cosmic Rays at Earth}. Elsevier.
  \item Smart, D. F., and Shea, M. A. (2005), A review of geomagnetic cutoff rigidities for cosmic ray measurements, \emph{Advances in Space Research}, 36, 2012--2020.
  \item NOAA/SWPC, GOES proton flux product documentation (pfu conventions).
  \item St\"ormer, C. (1955), \emph{The Polar Aurora}. Oxford University Press.
\end{itemize}

\end{document}
